Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-09-30 papers/paper1/component_model.tex
Marking phrases we should reconsider.

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
%% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
We then describe extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time.

@@

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
The scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions,  subject to the precondition using the value produced by the output action.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

The event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
The automaton that prescribes a binding is said to \emph{own} that binding.
The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
For example, the fully-qualified name of the $in?(s)$ action of Converter C in Figure~\ref{sys_model} is $[root].C.in?(s)$.
Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
The map of bindings for the event detection system in Figure~\ref{sys_model} is: 
\begin{displaymath}
\begin{split}
\{ ([root].out!(s), [root].C.in?(s)),\\
   ([root].C.out!(t), [root].D.in?(t)),\\
   ([root].D.detect!, [root].detected?) \}.
\end{split}
\end{displaymath}

\paragraph*{Binding rules}
To support name-based composition under the I/O automata model, the set of bindings must adhere to four essential rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., use of a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component, even if the latter is a child of the former as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph*{Concurrent execution}
Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $process1$ and $out!(s)$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata directly.
%% Consequently, we add the ability to deal with a dynamic set of automata and interactions.

@@

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and is used in the execution of every action.
A \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

@@

To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
To do this, we take the conjunction of the precondition of an output action with a predicate over the set of bindings $B$ (called a \emph{binding predicate}) to ensure that the output action is not executed until its set of implied automata is correct.
Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as was previously discussed.
Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
A system defined statically can be implemented as a dynamic system adding binding predicates and logic to create the required constellation of automata.

>>>>

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
%% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
We then describe \textbf{extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time}.

@@

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
The scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

The event detector system illustrated in Figure~\ref{sys_model} contains three component automata:  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

%% This is only true for static systems.  Dynamic system can do crazy things.
Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
The automaton that prescribes a binding is said to \emph{own} that binding.
The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
For example, the fully-qualified name of the $in?(s)$ action of Converter C in Figure~\ref{sys_model} is $[root].C.in?(s)$.
Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
The map of bindings for the event detection system in Figure~\ref{sys_model} is: 
\begin{displaymath}
\begin{split}
\{ ([root].out!(s), [root].C.in?(s)),\\
   ([root].C.out!(t), [root].D.in?(t)),\\
   ([root].D.detect!, [root].detected?) \}.
\end{split}
\end{displaymath}

\paragraph*{Binding rules}
To support name-based composition under the I/O automata model, the set of bindings must adhere to four essential rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., use of a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component, \textbf{even if the latter is a child of the former} as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph*{Concurrent execution}
Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $out!(s)$ and $process1$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata \textbf{directly}.
%% Consequently, we add the ability to deal with a dynamic set of automata and interactions.

@@

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and is used in the execution of every action.
A \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

@@

To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
To do this, we take the conjunction of the precondition of an output action with a predicate over the set of bindings $B$ (called a \emph{binding predicate}) to ensure that the output action is not executed until its set of implied automata is correct.
Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as was previously discussed.
Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
A system defined statically can be implemented as a dynamic system by adding binding predicates and logic to create the required constellation of automata.

### 2011-09-30 papers/paper1/design.tex
Marking phrases we should reconsider.

The controller encodes a \emph{policy} that selects the next local action.
For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
The run-time system can be initialized with different controllers to achieve different performance objectives.
Controllers can use multiple threads of execution to exploit the concurrency available in a system of automata.
The ioa++ framework is designed so that all native concurrency structures (e.g., threads, mutexes) are encapsulated in the dispatcher and controller.

@@

A more focused approached used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare actions to the scheduler after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

@@

However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
A simple technique for accomplishing this is to \emph{schedule} all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
However, this must be done carefully.
Not declaring an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action could be another potential mistake.

>>>>

The controller encodes a \emph{policy} that selects the next local action.
For the framework to be a valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
The run-time system can be initialized with different controllers to achieve different performance objectives.
Controllers can use multiple threads of execution to exploit the concurrency available in a system of automata.
The ioa++ framework is designed so that all native concurrency mechanisms (e.g., threads, mutexes) are encapsulated in the dispatcher and controller.

@@

A more focused approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare actions to the scheduler after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

@@

However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
A simple technique for accomplishing this is to \emph{schedule} all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
However, this must be done carefully.
Not declaring an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action \textbf{could be another potential mistake}.
